---
title: "Example visualizations for Demo"
author: "Andrew"
date: "15 July 2022"
output: html_document
---

This notebook is used to guide the medcomms data visualization demo
```{r setup, include=FALSE}
library(tidyverse)
library(here)
# Datasets
library(gapminder)
```



# Set theme

```{r}
# Set theme (very minimal)
theme_set(theme_bw())

theme_update(
  axis.line = element_line(colour = "black"),
  axis.ticks = element_line(colour = "black"),
  axis.text = element_text(color = "black"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.border = element_blank(),
  panel.background = element_blank(),
  plot.margin = margin(rep(15, 4)),
  strip.background = element_rect(colour = NA, fill = NA),
  strip.text = element_text(face = "bold", hjust = 0),
  text = element_text(size = 12)
)
```

# iris dataset
Here we will use the iris dataset to create a basic boxplot

```{r}
iris_boxplot <- ggplot(data = iris) +
  aes(x = Species, y = Sepal.Length, color = Species) +
  geom_boxplot() +
  geom_jitter(position = position_jitter(0.2)) +
  scale_y_continuous(breaks = seq(0, 8, by = 2), limits = c(0, 8), expand = c(0, 0)) +
  xlab(NULL) +
  ylab("Sepal length (cm)")

iris_boxplot

ggsave(here("output", "iris_boxplot.png"), iris_boxplot)
```




# Gapminder data

Here, we can examine the gapminder data (GDP per capita for year versus life expectancy for year 2007) using a more complex plot. Taken from here: https://bookdown.org/jbrophy115/bookdown-clinepi/vis.html

```{r}

# Focus on 2007 only 
data <- gapminder %>% 
  filter(year == 2007)

gapminder_graph <- ggplot(data, aes(x = gdpPercap, y = lifeExp)) +
    geom_point(aes(size = pop, color = continent), alpha = 0.5) +
    stat_smooth(formula = y ~ log10(x), se = FALSE, size = 0.5, color = "black", method = 'gam', linetype="dotted") +
    xlab("GDP per capita") +
    ylab("Life expectancy at birth") +
    scale_size_area(guide = "none", max_size = 15) +
    scale_x_continuous(labels = scales::dollar, breaks = seq(0, 50000, by = 10000), limits = c(0, 50000), expand = c(0, 0)) +
    scale_y_continuous(breaks = seq(0, 90, by = 10), limits = c(0, 90), expand = c(0, 0)) +
    scale_color_brewer(name = "", palette = "Dark2") 

gapminder_graph

ggsave(here("output", "gapminder_chart.png"), gapminder_graph)

```


# Optional: animated graph for gapminder data above (save as gif)
```{r}
library(gganimate)
# if you get a "Error in transform_path", you will need to install the 'transformr' library

# Here we use gganimate to look at 30 years worth of data as an animation
data_filtered_30yrs <- gapminder %>%
  filter(year > 1976)

animated_plot <- ggplot(data_filtered_30yrs, aes(x = gdpPercap, y = lifeExp)) +
    geom_point(aes(size = pop, color = continent), alpha = 0.5) +
    stat_smooth(formula = y ~ log10(x), se = FALSE, size = 0.5, color = "black", method = 'gam', linetype="dotted") +
    xlab("GDP per capita") +
    ylab("Life expectancy at birth") +
    scale_size_area(guide = "none", max_size = 15) +
    scale_x_continuous(labels = scales::dollar, breaks = seq(0, 60000, by = 10000), limits = c(0, 60000), expand = c(0, 0)) +
    scale_y_continuous(breaks = seq(0, 90, by = 10), limits = c(0, 90), expand = c(0, 0)) +
    scale_color_brewer(name = "", palette = "Dark2")  +
    # gganimate code
    ggtitle("{frame_time}") +
    transition_time(year) +
    ease_aes("linear") +
    enter_fade() +
    exit_fade()

# it might take a while to run!
animated_plot

# Save as gif
anim_save(here("output", "gapminder_animated.gif"), animated_plot)

```


